Distributional Trees and Forests for Circular Data

### 2019-10-06 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revisions and restructuring of section 'Evaluation Setting and Models Applied'; small changes in motivation for distributional trees and forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2789 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 386e7c5ecf000263b56206a789b1888806aa3e14

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
allowing for the mean direction or the dispersion to be modeled not only in separate models 
but also in a mixed model. However, applying these models has shown to be very challenging, 
mainly due to the complexity encountered in optimizing the log-likelihood function.
Moreover the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in [0, 2\pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative approach that can deal with these problems is to apply a tree-structured model.

@@

\subsection{Reference Models}
Along with distributional trees for circular data and forests three other models have been applied 
and their performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.
% TODO: references
To establish a distributional model of persistence for this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the exact same time points of the previous six days are considered. In order to account
for a stronger influence of observations closer to the prediction time point exponentially
increasing weights are included with the highest weight being assigned to the time point of the
previous day. For this selected set of observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.

@@

\subsection{Evaluation}
Five-fold cross validation (four years as learning data, one year as testing data) ...

Compare fitted models based on a circular version of CRPS ...

>>>>

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for the mean direction or the dispersion 
to be modeled not only in separate models but also in a mixed model. However, applying these models 
has shown to be very challenging, mainly due to the complexity encountered in optimizing the 
log-likelihood function.
Additionally, the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
parameters of the von Mises distribution, however, in an iterative optimization procedure
requiring well fitting initial values for the parameters that are recommended to be calculated 
in advance in two separate models. Using the obtained initial value the model parameters are 
estimated and improved stepwise until convergence. While providing a well elaborated theory 
this methodology and the contained optimization problem have shown to be very complex and 
bear the risk of not converging if not applied thoughtfully \textit{FIXME: citation}. %\citealp{}.
Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) is calculated and summed up over each observation contained in
the learning data and the corresponding predicted distribution parameters of each of the models.

To test whether distributional trees and forests can be applied as a reasonable alternative to already
existing approaches to modeling circular data, particularly in the field of meteorological forecasting,
three commonly used models are selected as reference models.
The exact settings of the tree and forest models as well as a specification of the reference models
is provided in the remainder of this section.

\paragraph{Distributional Tree}
For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
number of 2000 observations in each terminal node serving as a stopping criterion 
(argument \code{minbucket} of our implementation).

\paragraph{Distributional Forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points lead to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.
% TODO: references
To establish a distributional model of persistence for this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the exact same time points of the previous six days are considered. In order to account
for a stronger influence of observations closer to the prediction time point exponentially
increasing weights are included with the highest weight being assigned to the time point of the
previous day. For this selected set of observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

@@
